---
title: "What job does the cutter actually do?"
subtitle: "2024–2026 regular season"
author: "Henry Blaney"
output:
  html_document:
    theme: paper
    toc: false
    fig_width: 7.2
    fig_height: 4.2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
library(dplyr)
library(ggplot2)
library(knitr)

pct <- function(x) paste0(sprintf("%.1f", 100 * x), "%")
num <- function(x, d = 2) sprintf(paste0("%.", d, "f"), x)

p2024 <- readRDS("data/pitches_2024.rds")
p2025 <- readRDS("data/pitches_2025.rds")
p2026 <- readRDS("data/pitches_2026.rds")
pitches <- bind_rows(p2024, p2025, p2026)
rm(p2024, p2025, p2026)

pitches <- pitches |>
  mutate(
    is_take = description %in% c("ball", "blocked_ball", "called_strike"),
    is_swing = description %in% c(
      "foul", "foul_tip", "swinging_strike",
      "swinging_strike_blocked", "hit_into_play"
    ),
    is_called_strike = description == "called_strike",
    is_whiff = description %in% c("swinging_strike", "swinging_strike_blocked"),
    rv = delta_pitcher_run_exp,
    in_zone = zone %in% 1:9,
    situation = case_when(
      strikes == 2 ~ "two_strikes",
      balls > strikes ~ "behind",
      TRUE ~ "other"
    )
  )

keep <- c("FF", "SI", "FC", "SL", "ST", "CU", "CH")

league <- pitches |>
  filter(pitch_type %in% keep) |>
  group_by(year, pitch_type) |>
  summarise(
    n = n(),
    cs_per_take = sum(is_called_strike) / sum(is_take),
    whiff_per_swing = sum(is_whiff) / sum(is_swing),
    .groups = "drop"
  )

job_count <- pitches |>
  filter(pitch_type %in% c("FF", "FC", "SL")) |>
  group_by(pitch_type, situation) |>
  summarise(
    n = n(),
    cs_per_take = sum(is_called_strike) / sum(is_take),
    whiff_per_swing = sum(is_whiff) / sum(is_swing),
    rv_per_100 = 100 * mean(rv, na.rm = TRUE),
    .groups = "drop"
  )

cutters <- pitches |>
  filter(pitch_type == "FC") |>
  group_by(year, pitcher_id) |>
  summarise(
    n = n(),
    cs_per_take_pp = sum(is_called_strike) / sum(is_take),
    whiff_per_swing_pp = sum(is_whiff) / sum(is_swing),
    avg_run_value_per_100 = 100 * mean(rv, na.rm = TRUE),
    pct_two_strikes = mean(strikes == 2, na.rm = TRUE),
    zone_pp = mean(in_zone, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n >= 150)

cors <- list(
  cs_whiff = cor(cutters$cs_per_take_pp, cutters$whiff_per_swing_pp),
  cs_2k = cor(cutters$cs_per_take_pp, cutters$pct_two_strikes),
  whiff_2k = cor(cutters$whiff_per_swing_pp, cutters$pct_two_strikes),
  cs_zone = cor(cutters$cs_per_take_pp, cutters$zone_pp)
)

yoy_rates <- function(y1, y2) {
  inner_join(
    cutters |>
      filter(year == y1) |>
      select(pitcher_id, cs1 = cs_per_take_pp, wh1 = whiff_per_swing_pp,
             rv1 = avg_run_value_per_100, k1 = pct_two_strikes),
    cutters |>
      filter(year == y2) |>
      select(pitcher_id, cs2 = cs_per_take_pp, wh2 = whiff_per_swing_pp,
             rv2 = avg_run_value_per_100, k2 = pct_two_strikes),
    by = "pitcher_id"
  )
}
y45 <- yoy_rates(2024, 2025)
y56 <- yoy_rates(2025, 2026)

cutter_loc <- pitches |>
  filter(pitch_type == "FC", !is.na(plate_x)) |>
  semi_join(cutters, by = c("year", "pitcher_id")) |>
  mutate(
    matchup = if_else(pitch_hand == bat_side, "same", "opp"),
    glove_x = if_else(pitch_hand == "R", plate_x, -plate_x),
    side = if_else(glove_x >= 0, "glove", "arm"),
    loc_label = case_when(
      matchup == "same" & side == "glove" ~ "Same-handed, away (glove)",
      matchup == "same" & side == "arm" ~ "Same-handed, in (arm)",
      matchup == "opp" & side == "glove" ~ "Opposite-handed, in (glove)",
      TRUE ~ "Opposite-handed, away (arm)"
    )
  )

loc_2x2 <- cutter_loc |>
  group_by(matchup, side, loc_label) |>
  summarise(
    n = n(),
    cs_per_take = sum(is_called_strike) / sum(is_take),
    whiff_per_swing = sum(is_whiff) / sum(is_swing),
    rv_per_100 = 100 * mean(rv, na.rm = TRUE),
    .groups = "drop"
  )

cells <- cutter_loc |>
  group_by(year, pitcher_id, matchup, side) |>
  summarise(
    n = n(),
    cs = sum(is_called_strike) / sum(is_take),
    whiff = sum(is_whiff) / sum(is_swing),
    rv100 = 100 * mean(rv, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n >= 25)

paired_glove <- inner_join(
  cells |>
    filter(matchup == "same", side == "glove") |>
    select(year, pitcher_id, rv_s = rv100, cs_s = cs, wh_s = whiff),
  cells |>
    filter(matchup == "opp", side == "glove") |>
    select(year, pitcher_id, rv_o = rv100, cs_o = cs, wh_o = whiff),
  by = c("year", "pitcher_id")
)

tt_rv <- t.test(paired_glove$rv_s, paired_glove$rv_o, paired = TRUE)
tt_wh <- t.test(paired_glove$wh_s, paired_glove$wh_o, paired = TRUE)
tt_cs <- t.test(paired_glove$cs_s, paired_glove$cs_o, paired = TRUE)

med_cs <- median(cutters$cs_per_take_pp, na.rm = TRUE)
med_wh <- median(cutters$whiff_per_swing_pp, na.rm = TRUE)
cutters <- cutters |>
  mutate(
    cutter_type = case_when(
      cs_per_take_pp >= med_cs & whiff_per_swing_pp >= med_wh ~ "both",
      cs_per_take_pp >= med_cs & whiff_per_swing_pp < med_wh ~ "strike-leaning",
      cs_per_take_pp < med_cs & whiff_per_swing_pp >= med_wh ~ "whiff-leaning",
      TRUE ~ "neither"
    )
  )

stab <- function(y1, y2) {
  a <- cutters |> filter(year == y1) |> select(pitcher_id, t1 = cutter_type)
  b <- cutters |> filter(year == y2) |> select(pitcher_id, t2 = cutter_type)
  d <- inner_join(a, b, by = "pitcher_id")
  mean(d$t1 == d$t2)
}

type_palette <- c(
  FF = "#1b4f72", SI = "#5dade2", FC = "#c0392b",
  SL = "#117a65", ST = "#1e8449", CU = "#6c3483", CH = "#b7950b"
)
```

## Question

I wanted to know whether cutters get called strikes or whiffs, and whether pitchers throw them in counts that do not match that. The example I had in mind was finishing an at-bat with a cutter that only gets taken for strikes.

Run value is `delta_pitcher_run_exp` (pitcher side; I did not flip it). Called strike means they took it and it was a strike. Whiff means swinging strike, not foul tip. Pitcher-years need 150 cutters.

## Data

Regular-season Statcast, 2024–2026. About 169,000 cutters. `r nrow(cutters)` pitcher-years after the 150 cutoff. I have two year-to-year comparisons from that. I did not pull 2022 or 2023.

## Called strikes and whiffs

```{r fig-cs, fig.cap="Called-strike rate on takes, by pitch type."}
ggplot(league, aes(year, cs_per_take, color = pitch_type)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_color_manual(values = type_palette, name = "Pitch") +
  scale_y_continuous(labels = function(x) paste0(round(100 * x), "%")) +
  labs(title = "Called strikes per take", x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r fig-wh, fig.cap="Whiff rate on swings, by pitch type."}
ggplot(league, aes(year, whiff_per_swing, color = pitch_type)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_color_manual(values = type_palette, name = "Pitch") +
  scale_y_continuous(labels = function(x) paste0(round(100 * x), "%")) +
  labs(title = "Whiffs per swing", x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

On takes, the cutter’s called-strike rate is about 33%, next to four-seams. On swings it whiffs about 21%, also next to four-seams. Sliders are lower on called strikes and much higher on whiffs.

```{r job-table}
job_count |>
  filter(situation %in% c("behind", "two_strikes")) |>
  mutate(
    Pitch = recode(pitch_type, FC = "Cutter", FF = "Four-seam", SL = "Slider"),
    Count = recode(situation, behind = "Behind", two_strikes = "Two strikes")
  ) |>
  transmute(
    Pitch, Count,
    N = format(n, big.mark = ","),
    `CS / take` = pct(cs_per_take),
    `Whiff / swing` = pct(whiff_per_swing),
    `RV / 100` = num(rv_per_100)
  ) |>
  kable(caption = "Behind vs two strikes. 2024–2026 pooled.")
```

With two strikes the cutter still whiffs 20% of swings, same as when the pitcher is behind. The slider in that count is 29% whiffs and +0.38 runs per 100. The cutter is about zero. The four-seam is −0.21.

## Usage and labels

On the 150-plus pitcher-years, called-strike rate and whiff rate correlate **`r num(cors$cs_whiff)`**. Called-strike rate and zone rate correlate **`r num(cors$cs_zone)`**. Most of the “strike” split is location.

I thought the high called-strike guys might be using it too much with two strikes. They are not. Higher called-strike rate goes with less two-strike use (`r` = **`r num(cors$cs_2k)`**). Higher whiff rate goes with more (`r` = **`r num(cors$whiff_2k)`**).

Two-strike share year to year is **`r num(cor(y45$k1, y45$k2))`** then **`r num(cor(y56$k1, y56$k2))`**. Last year’s whiff rate predicts this year’s two-strike share. Called-strike rate is around 0.5 year to year. Run value is about zero. If I split guys at the median into strike-leaning and whiff-leaning, about half keep the same label next year (**`r pct(stab(2024, 2025))`**, **`r pct(stab(2025, 2026))`**), and the label does not predict next year’s run value.

I also built a same-year table of those labels by count. Strike-leaning looked bad with two strikes. That uses this year’s rates to make the label and this year’s pitches to score it, so I am leaving it out. `select(type)` after baseballr does not keep your column; I used `transmute(cutter_type = type)`.

## Location

Glove-side is the way the cutter breaks. For a right-hander that is first base: away from a right-handed hitter, in on a left-handed hitter. I had that backwards at first and said same-handed glove was in on the hands. Same-handed glove is away. Opposite-handed glove is in. Arm-side is the other half (including backdoor to a lefty). The pitch still cuts glove-side. This is just where it crosses the plate.

```{r fig-loc, fig.cap="Run value per 100 cutters by matchup and plate half."}
loc_2x2 |>
  mutate(cell = recode(loc_label,
    "Same-handed, away (glove)" = "Same / away",
    "Same-handed, in (arm)" = "Same / in",
    "Opposite-handed, in (glove)" = "Opp / in",
    "Opposite-handed, away (arm)" = "Opp / away"
  )) |>
  ggplot(aes(cell, rv_per_100, fill = matchup)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  scale_fill_manual(values = c(opp = "#7f8c8d", same = "#c0392b"), guide = "none") +
  labs(
    title = "Cutter run value per 100",
    x = NULL, y = "Pitcher RV / 100"
  ) +
  theme_minimal(base_size = 12)
```

```{r loc-table}
loc_2x2 |>
  transmute(
    Cell = loc_label,
    N = format(n, big.mark = ","),
    `CS / take` = pct(cs_per_take),
    `Whiff / swing` = pct(whiff_per_swing),
    `RV / 100` = num(rv_per_100)
  ) |>
  kable(caption = "Pitcher-years with 150 or more cutters. Glove is the half the pitch breaks toward.")
```

Same-handed away has the best run value. Opposite-handed in has the worst. Arm-side rows have the highest called-strike rates; those pitches are in the zone more often.

This is the whole glove-side half, not just a pitch on the hands. Whiffs and run value are on the away, same-handed side.

Same pitcher, both glove-side versions, 25 or more pitches in each cell (`r nrow(paired_glove)` pitcher-years):

- Away / same-handed: **`r num(mean(paired_glove$rv_s))`** RV / 100  
- In / opposite-handed: **`r num(mean(paired_glove$rv_o))`**  
- Difference **`r num(unname(tt_rv$estimate))`** per 100 (paired t, p = `r format(tt_rv$p.value, digits = 2)`)  
- Whiffs **`r pct(unname(tt_wh$estimate))`** higher on the away one  
- Called-strike rate is not higher on the in version

## So what?

Called-strike and whiff rates sit near the four-seam. With two strikes the cutter’s whiff rate does not go up; the slider row is still +0.38 per 100. Extra value in this file is glove-side and away from the same-handed hitter. One season of rates is not enough to relabel a pitcher and change his counts. Two-strike usage already follows whiff rate.

## Limits

The 150 cutoff drops light cutter use. 2026 called-strike rates can differ from 2024–2025 because of ABS / the plate-location change; that is why the first graphs are by year. Median types do not hold next year. The location table is a half of the plate. No sequencing. Public Statcast.

